A tutorial on spatiotemporal partially observed Markov process models via the R package spatPomp

### 2024-10-25 ms.Rnw
tentatively finished ben bolker suggestions

The initial state of the latent process, $\vec{X}_0=\vec{X}(t_0;\vectheta)$, is a draw from the initial distribution, $f_{\vec{X}_0}(\vec{x}_0\giventh\vectheta)$. 
If the initial conditions are known, there is no dependence on $\vectheta$.
Alternatively, there may be components of the $\vectheta$ having the sole function of specifying $\vec{X}_0$.
These components are called \emph{initial value parameters} (IVPs).
By contrast, parameters involved in the transition density or measurement density are called \emph{regular parameters} (RPs).
This gives rise to a decomposition of the parameter vector, $\theta=(\theta_{\RP},\theta_{\IVP})$.
We may specify $f_{\vec{X}_0}(\vec{x}_0\giventh \theta_{\RP},\theta_{\IVP})$ to be a point mass at $\theta_{\IVP}$, in which case $\theta_{\IVP}$ exactly corresponds to $\vec{X}_0$.
The \code{bm10} model has this structure, and the initialization can be tested by \code{rinit(bm10)}.
The measles model of \cref{sec:measles:spatPomp} specifies $\vec{X}_0$ as a deterministic function of $\theta_{\IVP}$, but not an identity map since it is convenient to describe latent states as counts and the corresponding IVPs as proportions.

@@

measles_plot_data <- plot(measles6,log=TRUE) +
  labs(y="") +
  theme(axis.text.y=element_blank(),
    plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  ylim(0,4) +
  scale_x_continuous(breaks = c(1950,1960))

measles_plot_sim <- plot(measles_sim,log=TRUE) +
  theme(axis.title.y=element_blank(),
    plot.margin=unit(c(0, 0, 0, 0), "cm")) +
  ylim(0,4) +
  scale_x_continuous(breaks = c(1950,1960))

>>>>

The initial state of the latent process, $\vec{X}_0=\vec{X}(t_0;\vectheta)$, is a draw from the initial distribution, $f_{\vec{X}_0}(\vec{x}_0\giventh\vectheta)$. 
If the initial conditions are known, there is no dependence on $\vectheta$.
Alternatively, there may be components of the $\vectheta$ having the sole function of specifying $\vec{X}_0$.
These components are called \emph{initial value parameters} (IVPs).
By contrast, parameters involved in the transition density or measurement density are called \emph{regular parameters} (RPs).
This gives rise to a decomposition of the parameter vector, $\theta=(\theta_{\RP},\theta_{\IVP})$.
We may specify $f_{\vec{X}_0}(\vec{x}_0\giventh \theta_{\RP},\theta_{\IVP})$ to be a point mass at $\theta_{\IVP}$, in which case $\theta_{\IVP}$ exactly corresponds to $\vec{X}_0$.
The \code{bm10} model has this structure, and the initialization can be tested by \code{rinit(bm10)}.
The measles model of \cref{sec:measles:spatPomp} specifies $\vec{X}_0$ as a deterministic function of $\theta_{\IVP}$, but not an identity map since it is convenient to describe latent states as counts and the corresponding IVPs as proportions. 
The RPs are provided to \code{rinit} so they can also participate in the initialization.

For a stationary model, the initial state may be set at the stationary distribution by specifying $t_0$ sufficiently remote from $t_1$ to allow the system to equilibrate.
\citet{stocks20} provides an example of this using \pkg{pomp}.
When following this approach, the model is insensitive to the choice of $f_{\vec{X}_0}(\vec{x}_0\giventh\vectheta)$.

@@

measles_plot_data <- plot(measles6,log=TRUE) +
  theme(axis.text.y=element_blank(),
    plot.margin = unit(c(0, 0, 0, 0), "cm")) +
  scale_y_log10(limits=c(1,10^4)) +
  scale_x_continuous(breaks = c(1950,1960))

measles_plot_sim <- plot(measles_sim,log=TRUE) +
  theme(axis.title.y=element_blank(),
    plot.margin=unit(c(0, 0, 0, 0), "cm")) +
  scale_y_log10(limits=c(1,10^4)) +
  scale_x_continuous(breaks = c(1950,1960))
